\documentclass{article}

% Input packages & formatting
\input{template/packages}
\input{template/formatting}
\input{template/version}
\renewcommand{\cleartooddpage}[1][]{\ignorespaces} % single side
\newcommand{\caret}{$^\wedge$}

\title{\Huge Tin: A Tcl Package Manager\\\small Version \version}
\author{Alex Baker\\\small\url{https://github.com/ambaker1/Tin}}
\date{\small\today}
\begin{document}
\maketitle
\begin{abstract}
Tin is a package installer for Tcl. 
With Tin, you can easily install Tcl packages directly from GitHub.
\end{abstract}
\clearpage
\section{The Official Tin List}
The Official Tin List is a database of Tcl packages that can be installed directly from GitHub. \\
As of Tin version \version, these are the packages in the Official Tin List:
\input{template/TinList.tex}
\clearpage
\section{Adding to the Tin List}
Tin installs packages based on entries in the ``Tin List'', which can be added to in the current session with the commands \cmdlink{tin add}.
\begin{syntax}
\command{tin add} <-tin> \$name \$version \$repo \$tag \$file
\end{syntax}
\begin{args}
-tin & Option to add an entry to the Tin List. Default. \\
\$name & Package name. \\
\$version & Package version. \\
\$repo & Github repository URL. \\
\$tag & Github release tag for version.  \\
\$file & Installer file path in repo. 
\end{args}
\begin{example}{Adding entries to the Tin List}
\begin{lstlisting}
tin add foo 1.0 https://github.com/username/foo v1.0 install_foo.tcl
tin add foo 1.1 https://github.com/username/foo v1.1 install_foo.tcl
tin add foo 1.2 https://github.com/username/foo v1.2 install_foo.tcl
tin add foo 1.2.1 https://github.com/username/foo v1.2.1 install_foo.tcl
tin add foo 1.2.2 https://github.com/username/foo v1.2.2 install_foo.tcl
tin add foo 2a0 https://github.com/username/foo v2a0 install_foo.tcl
tin add foo 2.0 https://github.com/username/foo v2.0 install_foo.tcl
tin add foo 2.0.1 https://github.com/username/foo v2.0.1 install_foo.tcl
\end{lstlisting}
\end{example}
Clearly, this is a bit verbose... and repetitive...
\clearpage
\section{Auto-Tin Configuration}
To streamline the process of populating the Tin List, simply use the command \cmdlink{tin add -auto} to specify an Auto-Tin configuration.
An Auto-Tin configuration specifies a template for Tin entries, and this template is used in conjunction with GitHub release tags to automatically populate the Tin List.
\begin{syntax}
\command{tin add -auto} \$name \$repo \$file <{}<-exact> \$version> <\$reqs ...>
\end{syntax}
\begin{args}
-auto & Option to add an Auto-Tin configuration to the Tin List. \\
\$name & Package name. \\
\$repo & Github repository URL. \\
\$file & Installer file path in repo. \\
-exact & Option to specify exact version. \\
\$version & Package version. \\
\$reqs ... & Package version requirements, mutually exclusive with -exact option. 
\end{args}
\subsection{Auto-Tin Tag Pattern}
In order for an Auto-Tin configuration to work, the GitHub repository must have release tags corresponding directly with the package versions, such as ``v1.2.3''.

To be specific, version release tags must match the Auto-Tin tag pattern:
\begin{lstlisting}[language=verbatim]
^v(0|[1-9]\d*)(\.(0|[1-9]\d*))*([ab](0|[1-9]\d*)(\.(0|[1-9]\d*))*)?$
\end{lstlisting}
Note that this is not the same as ``SemVer'' (\url{https://semver.org/}), which is the standard version number format. 
This is because the Auto-Tin tag pattern specifically matches all version numbers compatible with Tcl, with a prefix of ``v''. 
Most notably, the format for alpha and beta versions is different in Tcl, where ``a'' and ``b'' replace one of the periods (e.g. ``v2a0''), and effectively represent ``.-2.'' and ``.-1.'', respectively.
\clearpage
\subsection{Fetching and Auto-Fetching}
If a package is configured as an Auto-Tin package, the Tin List can be automatically populated with versions available for installation with the command \cmdlink{tin fetch}. 
If the -all option is specified, it will return a dictionary of package names and versions added.
Otherwise, it will simply return a list of the versions added for the package. 
\begin{syntax}
\command{tin fetch} \$name <\$pattern> \\
tin fetch -all <\$names>
\end{syntax}
\begin{args}
\$name & Package name. \\
\$pattern & Version number glob pattern for \textit{git ls-remote}. Default ``\texttt{*}'', or all versions. \\
-all & Option to fetch all available versions. \\
\$names & List of package names. Default all Auto-Tin packages.
\end{args}
Additionally, by default, Tin auto-fetches on the fly when it cannot find Tin entries satisfying version requirements. 
This can be toggled on/off with \cmdlink{tin auto}, which also returns the current auto-fetch status.
\begin{syntax}
\command{tin auto} <\$toggle>
\end{syntax}
\begin{args}
\$toggle & Boolean, whether to auto-fetch. Default true.
\end{args}
\begin{example}{Adding an Auto-Tin configuration and fetching Tin entries}
\begin{lstlisting}
tin add -auto foo https://github.com/username/foo install_foo.tcl
tin fetch foo; # fetches all entries from GitHub
\end{lstlisting}
\end{example}

\clearpage
\section{Removing from the Tin List}
Both Tin entries and Auto-Tin configurations can be removed from the Tin List in a session with the command \cmdlink{tin remove}. 
\begin{syntax}
\command{tin remove} <-tin> \$name <\$version> <\$repo> \\
tin remove -auto \$name <\$repo> <\$file> \\
tin remove -all <\$names>
\end{syntax}
\begin{args}
-tin & Option to remove a Tin entry from the Tin List. Default. \\
-auto & Option to remove an Auto-Tin configuration from the Tin List. \\
-all & Option to remove both Tin entries and Auto-Tin configurations. \\
\$name & Package name. \\
\$names & List of package names (default all packages). \\
\$version & Package version to remove (optional, default all versions). \\
\$repo & Repository to remove (optional, default all repositories). \\
\$file & Installer file path to remove (optional, default all installer files).
\end{args}
\subsection{Clearing the Tin List}
The command \cmdlink{tin clear} removes all entries and Auto-Tin configurations from the Tin List, same as \textit{tin remove -all}.
\begin{syntax}
\command{tin clear}
\end{syntax}


\clearpage
\section{Saving and Resetting the Tin List}
The state of the Tin List can be saved for future sessions with \cmdlink{tin save}, and reset to default or factory settings with \cmdlink{tin reset}. 
Note that \cmdlink{tin save} saves to a hidden user-config file located in the user's home directory.
\begin{syntax}
\command{tin save}
\end{syntax}
\begin{syntax}
\command{tin reset} <-soft> \\
tin reset -hard
\end{syntax}
\begin{args}
-soft & Option to reset to user settings (default). \\
-hard & Option to reset to factory settings.
\end{args}

\begin{example}{Saving changes to the Tin List}
\begin{lstlisting}
tin reset -hard
tin add foo 1.0 https://github.com/username/foo v1.0 install_foo.tcl
tin save
\end{lstlisting}
\tcblower

\textit{"$\sim$/.tinlist.tcl" :}
\begin{lstlisting}
tin add foo 1.0 https://github.com/username/foo v1.0 install_foo.tcl
\end{lstlisting}
\end{example}


\clearpage
\section{Available Packages}
The available packages in the Tin List can be queried with the command \cmdlink{tin packages}.
\begin{syntax}
\command{tin packages} <\$pattern> \\
tin packages -tin <\$pattern> \\
tin packages -auto <\$pattern>
\end{syntax}
\begin{args}
-tin & Option to search only for Tin packages. \\
-auto & Option to search only for Auto-Tin packages. \\
\$pattern & Optional ``glob'' pattern, default ``\texttt{*}'', or all packages.
\end{args}
\begin{example}{Getting available packages}
\begin{lstlisting}
set allPackages [tin packages]
set loadedPackages [tin packages -tin]
set autoPackages [tin packages -auto]
\end{lstlisting}
\end{example}
\subsection{Available Package Versions}
A list of available versions for a Tin package that satisfy version requirements can be queried with the command \cmdlink{tin versions}.
Similarly, the command \cmdlink{tin available} returns the version that would be installed with \cmdlink{tin install}, or blank if none can be found.
\begin{syntax}
\command{tin versions} \$name <{}<-exact> \$version> <\$reqs ...>
\end{syntax}
\begin{syntax}
\command{tin available} \$name <{}<-exact> \$version> <\$reqs ...>
\end{syntax}
\begin{args}
\$name & Package name. \\
-exact & Option to specify exact version. \\
\$version & Package version. \\
\$reqs ... & Package version requirements, mutually exclusive with -exact option.
\end{args}
\clearpage
\section{Accessing the Tin List}
The command \cmdlink{tin get} queries basic information about Tin, and returns blank if the requested entry does exist. 
Similar to \cmdlink{tin add} and \cmdlink{tin remove}, it has two forms, one for querying Tin packages and one for querying Auto-Tin packages. 
Returns a dictionary associated with the supplied arguments.
\begin{syntax}
\command{tin get} <-tin> \$name <\$version> <\$repo> \\
tin get -auto \$name <\$repo> <\$file>
\end{syntax}
\begin{args}
-tin & Option to query Tin packages (default). \\
-auto & Option to query Auto-Tin packages. \\
\$name & Package name. \\
\$version & Package version.  \\
\$repo & Github repository URL. \\
\$file & Installer file path in repo.
\end{args}
\begin{example}{Getting info from the Tin List}
\begin{lstlisting}
package require tin
tin add foo 1.0 https://github.com/username/foo v1.0 install_foo.tcl
puts [tin get foo]
\end{lstlisting}
\tcblower
\begin{lstlisting}
1.0 {https://github.com/username/foo {v1.0 install_foo.tcl}}
\end{lstlisting}
\end{example}

\clearpage
\section{Installing and Uninstalling Packages}
The command \cmdlink{tin install} installs packages directly from GitHub, and returns the version installed.

The command \cmdlink{tin depend} installs packages only if they are not installed, and returns the version number installed (useful for dependencies in installation scripts).

The command \cmdlink{tin uninstall} uninstalls packages (as long as they are in the Tin List), and returns blank if successful, or error if it was unsuccessful in uninstalling the package. 

\begin{syntax}
\command{tin install} \$name <{}<-exact> \$version> <\$reqs ...>
\end{syntax}
\begin{syntax}
\command{tin depend} \$name <{}<-exact> \$version> <\$reqs ...>
\end{syntax}
\begin{syntax}
\command{tin uninstall} \$name <{}<-exact> \$version> <\$reqs ...>
\end{syntax}
\begin{args}
\$name & Package name. \\
-exact & Option to specify exact version. \\
\$version & Package version. \\
\$reqs ... & Package version requirements, mutually exclusive with -exact option.
\end{args}

By default, uninstalling a package simply deletes the library folder associated with the package.
However, if a ``pkgUninstall.tcl'' file is located within the package folder, it will run that file instead, with the variable \texttt{\$dir} set to the package library folder path, similar to how ``pkgIndex.tcl'' files work.

\begin{example}{Complex uninstall file ``pkgUninstall.tcl''}
\begin{lstlisting}
set bindir [file dirname [info nameofexecutable]]
file delete [file join $bindir foo.bat]; # delete file in the bin directory
file delete -force $dir; # Clean up package
\end{lstlisting}
\end{example}
\clearpage
\section{Upgrading Packages}
Upgrading packages involves first installing the new version with \cmdlink{tin install}, then uninstalling the old version with \cmdlink{tin uninstall}.
The command \cmdlink{tin check} returns an upgrade list of available minor and patch upgrades for packages.
The command \cmdlink{tin upgrade} checks for available upgrades with \cmdlink{tin check}, and then upgrades the packages, returning the upgrade list. 
If there is no upgrade available, the upgrade list will be empty. 
If the -all option is specified, the format of the upgrade list is ``\texttt{name \{old new ...\} ...}''. 
Otherwise, the format of the upgrade list is ``\texttt{old new}''.
\begin{syntax}
\command{tin check} \$name <{}<-exact> \$version> <\$reqs ...> \\
tin check -all <\$names> 
\end{syntax}
\begin{syntax}
\command{tin upgrade} \$name <{}<-exact> \$version> <\$reqs ...> \\
tin upgrade -all <\$names> 
\end{syntax}
\begin{args}
-all & Option to look for upgrades in all installed major versions of the packages. \\
\$names & Package names. Default searches all packages in the Tin List.\\
\$name & Package name. \\
-exact & Option to specify exact version. \\
\$version & Package version. \\
\$reqs ... & Package version requirements, mutually exclusive with -exact option.
\end{args}
\begin{example}{Upgrading Tin}
\begin{lstlisting}
# Upgrade Tin
package require tin
tin upgrade tin
\end{lstlisting}
\end{example}

\clearpage
\section{Loading and Importing Packages}
The command \cmdlink{tin require} is similar to the Tcl command \textit{package require}, but with the added feature that if the package is missing, it will try to install it with \cmdlink{tin install}.

The command \cmdlink{tin import} additionally handles most use-cases of \textit{namespace import}. 
Both \cmdlink{tin require} and \cmdlink{tin import} return the version number of the package imported.
\begin{syntax}
\command{tin require} \$name <{}<-exact> \$version> <\$reqs ...>
\end{syntax}
\begin{args}
\$name & Package name. \\
-exact & Option to specify exact version. \\
\$version & Package version. \\
\$reqs ... & Package version requirements, mutually exclusive with -exact option.
\end{args}
\begin{syntax}
\command{tin import} <-force> <\$patterns from> \$name <{}<-exact> \$version> <\$reqs ...> <as \$ns>
\end{syntax}
\begin{args}
-force & Option to overwrite existing commands. \\
\$patterns & Commands to import, or ``glob'' patterns, default ``\texttt{*}'', or all commands. \\
\$name & Package name. \\
-exact & Option to specify exact version. \\
\$version & Package version. \\
\$reqs ... & Package version requirements, mutually exclusive with -exact option. \\
\$ns & Namespace to import into. Default global namespace, or ``\texttt{::}''.
\end{args}

\begin{example}{Importing all commands package ``foo''}
\begin{lstlisting}
package require tin
tin import foo 1.0
\end{lstlisting}
\end{example}
\clearpage
\section{Generic Package Utilities}
Tin works alongside the Tcl \textit{package} commands, and provides a few package utility commands that do not interface with the Tin List.
\subsection{Check if a Package is Installed}
The command \cmdlink{tin installed} returns the package version number that would be loaded with \textit{package require}, or blank if the package is not installed. 
If there is no package version in the Tcl package database satisfying the requirements, it will call the \cmdlink{package unknown} script to load \textit{package ifneeded} statements from ``pkgIndex.tcl'' files, just like what \textit{package require} does, but without loading the package.
\begin{syntax}
\command{tin installed} \$name <{}<-exact> \$version> <\$reqs ...>
\end{syntax}
\begin{args}
\$name & Package name. \\
-exact & Option to specify exact version. \\
\$version & Package version. \\
\$reqs ... & Package version requirements, mutually exclusive with -exact option.
\end{args}
\subsection{Unload a Package/Namespace}
The command \cmdlink{tin forget} is short-hand for both \textit{package forget} and \textit{namespace delete}, but it will not throw an error if there is no namespace corresponding with the package name.
It is especially useful for reloading packages within an instance of Tcl.
\begin{syntax}
\command{tin forget} \$name ...
\end{syntax}
\begin{args}
\$name & Package name. 
\end{args}
\begin{example}{Loading and reloading a package}
\begin{lstlisting}
package require foo
tin forget foo
package require foo
\end{lstlisting}
\end{example}
\clearpage
\section{Utilities for Package Development}
In addition to commands for installing and loading packages, Tin provides a few commands intended to help in writing installation and build files for your packages.
\subsection{Creating Package Directories}
The command \cmdlink{tin mkdir} creates a library directory to install a package in, with a normalized naming convention that allows it to be uninstalled easily with \cmdlink{tin uninstall}. 
\begin{syntax}
\command{tin mkdir} <-force> <\$basedir> \$name \$version
\end{syntax}
\begin{args}
-force & Option to create fresh library directory (deletes existing folder). \\
\$basedir & Base directory, default one folder up from the Tcl library folder. \\
\$name & Package name. \\
\$version & Package version.
\end{args}

See the example installation file for a package ``foo'' that requires the package ``bar 1.2'', and installs in library folder ``foo-1.0''.
\begin{example}{Example file ``install\textunderscore{}foo.tcl''}
\begin{lstlisting}
package require tin
tin depend bar 1.2
set dir [tin mkdir -force foo 1.0]
file copy README.md $dir
file copy LICENSE $dir
file copy lib/bar.tcl $dir
file copy lib/pkgIndex.pdf $dir
\end{lstlisting}
\end{example}

\clearpage
\subsection{Building Library Files from Source with Configuration Variable Substitution}
The command \cmdlink{tin bake} takes an input text file, and writes an output text file after substitution of configuration variables such as \texttt{@VERSION@}.
This is especially helpful for ensuring that the package version is consistent across the entire project.
If a source directory is used as input, it will batch bake all ``.tin'' files. 

\begin{syntax}
\command{tin bake} \$src \$target \$config \\
tin bake \$src \$target \$varName \$value ...
\end{syntax}
\begin{args}
\$src & Source file, or directory with ``.tin'' files. \\
\$target & Target file, or directory to write ``.tcl'' files to. \\
\$config & Dictionary of config variable names and values. Config variables must be uppercase alphanumeric. \\
\$varName \$value ... & Config variable names and values. Mutually exclusive with \texttt{\$config}.
\end{args}

See below for an example of how \cmdlink{tin bake} can be used to automatically update a ``pkgIndex.tcl'' file:

\begin{example}{Building a ``pkgIndex.tcl'' file}
\begin{lstlisting}
package require tin
tin bake pkgIndex.tin pkgIndex.tcl {VERSION 1.0}
\end{lstlisting}
\tcblower

\textit{"pkgIndex.tin" :}
\begin{lstlisting}
package ifneeded foo @VERSION@ [list source [file join $dir foo.tcl]]
\end{lstlisting}
\textit{"pkgIndex.tcl" :}
\begin{lstlisting}
package ifneeded foo 1.0 [list source [file join $dir foo.tcl]]
\end{lstlisting}
\end{example}
\end{document}



